---
title: "Final Report"
author: "Dallan Gardner, Marcie Mortensen, and Joseph Holmes"
format: html
---


```{r, Setup, include = FALSE}
library(tidyverse)
library(glmnet)
library(bestglm)
library(ggfortify)
library(car)   
library(corrplot)  
library(GGally)
library(patchwork)
```

```{r, Import, include = FALSE}
# Read in data and move response (Score) to last column for later model selection
happiness <- read.csv("Happiness.csv") |>
  select(-Score, everything(), -Overall.rank, -Country.or.region, Score)
```

```{r, Scatter Matrix, fig.align='center', echo = FALSE, message = FALSE}
# Multicollinearity check
ggpairs(happiness)
```

```{r, Correlation Plots, echo = FALSE}
corrplot(cor(happiness))
```

```{r, Model Selection, echo = FALSE, results = 'hide'}
base.model <- lm(Score ~ 1, happiness)
full.model <- lm(Score ~ ., happiness)

# Best Subsets BIC, AIC, and Cross Validation
bestglm(happiness, IC = "BIC")
bestglm(happiness, IC = "AIC")
bestglm(happiness, IC = "CV")

# AIC Forward Selection
step.f.aic <- step(base.model, 
                   direction = "forward",
                   k = 2,
                   scope = list(lower = base.model, upper = full.model))

# AIC Backward Selection
step.b.aic <- step(full.model, 
                   direction = "backward",
                   k = 2,
                   scope = list(lower = base.model, upper = full.model))

# BIC Forward Selection
step.f.bic <- step(base.model, 
                   direction = "forward",
                   k = log(nrow(happiness)),
                   scope = list(lower = base.model, upper = full.model))

# BIC Backward Selection
step.b.bic <- step(full.model, 
                   direction = "backward",
                   k = log(nrow(happiness)),
                   scope = list(lower = base.model, upper = full.model))

# Sequential Replacement Selection (AIC)
seq.repl.aic <- step(base.model, 
                   direction = "both",
                   k = 2,
                   scope = list(lower = base.model, upper = full.model))

# Sequential Replacement Selection (BIC)
seq.repl.bic <- step(base.model, 
                   direction = "both",
                   k = log(nrow(happiness)),
                   scope = list(lower = base.model, upper = full.model))

# LASSO Regression
happiness_x <- as.matrix(happiness[, 1:6]) # predictors
happiness_y <- happiness[, 7] # response

happiness_lasso_cv <- cv.glmnet(x = happiness_x,
                          y = happiness_y, 
                          type.measure = "mse", 
                          alpha = 1)

happiness_lasso_cv$lambda.1se
coef(happiness_lasso_cv, s = "lambda.1se")

# Ridge Regression
happiness_ridge_cv <- cv.glmnet(x = happiness_x,
                          y = happiness_y, 
                          type.measure = "mse", 
                          alpha = 0)

happiness_ridge_cv$lambda.1se
coef(happiness_ridge_cv, s = "lambda.1se")
```

```{r, Selection Summaries}
summary(step.f.aic)
summary(step.b.aic)
summary(step.f.bic)
summary(step.b.bic)
summary(seq.repl.aic)
summary(seq.repl.bic)
```

Other than best subset, we'll use BIC for the other selection methods.

Variable                        | Best Subset (BIC) | Best Subset (AIC) | Best Subset (CV) | Forward | Backward | Sequential Replacement | LASSO     | 
------------------------------- | ----------------- | ----------------- | ---------------- | --------| -------- | ---------------------- | --------- |
GDP.per.capita                  | 0.8105494         | 0.7454527         | 0.8105494        | 0.8105  | 0.8105   | 0.8105                 | 0.7029658 |
Social.support                  | 1.0166104         | 1.1180315         | 1.0166104        | 1.0166  | 1.0166   | 1.0166                 | 0.9307674 |
Healthy.life.expectancy         | 1.1414486         | 1.0840162         | 1.1414486        | 1.1414  | 1.1414   | 1.1414                 | 0.9661429 |
Freedom.to.make.life.choices    | 1.8458157         | 1.5340094         | 1.8458157        | 1.8458  | 1.8458   | 1.8458                 | 1.2050668 |
Generosity                      | .                 | .                 | .                | .       | .        | .                      | .         |
Perceptions.of.corruption       | .                 | 1.1175531         | .                | .       | .        | .                      | 0.2618378 |




Generosity was not selected in any process, so we can confidently remove that. Perception of corruption was selected in the AIC best subsets and LASSO selection, but nothing else. If we stick with BIC, we should probably remove this one too.

```{r, Create Model}
happiness <- happiness |>
  select(-Generosity, -Perceptions.of.corruption)

fit.happiness <- lm(Score ~ ., happiness)
happiness$Residuals <- fit.happiness$residuals
happiness$'Fitted Values' <- fit.happiness$fitted.values

summary(fit.happiness)
```

```{r, Linearity}
# Xs and Y are linear

# Partial Regression Plots
avPlots(fit.happiness)

# Residual by Predictor Plots
resid_vs_gdp <- ggplot(happiness, aes(x = GDP.per.capita, y = Residuals)) +
  geom_point() +
  theme_bw() +
  theme(aspect.ratio = 1)

resid_vs_social <- ggplot(happiness, aes(x = Social.support, y = Residuals)) +
  geom_point() +
  theme_bw() +
  theme(aspect.ratio = 1)

resid_vs_healthy <- ggplot(happiness, aes(x = Healthy.life.expectancy, y = Residuals)) +
  geom_point() +
  theme_bw() +
  theme(aspect.ratio = 1)

resid_vs_freedom <- ggplot(happiness, aes(x = Freedom.to.make.life.choices, y = Residuals)) +
  geom_point() +
  theme_bw() +
  theme(aspect.ratio = 1)

(resid_vs_gdp | resid_vs_social) / (resid_vs_healthy | resid_vs_freedom)
```

Linearity looks good.

# Check Independence Here

```{r, Normal Residuals}
# Errors are normally distributed
ggplot(happiness, aes(x = Score)) +
  geom_histogram(fill = 'blue', alpha = 0.5, color = 'black') +
  labs(
    title = "Residuals Histogram",
    y = "Count",
    x = "Happiness Score"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))


ggplot(happiness, aes(sample = Residuals)) +
  stat_qq() +
  stat_qq_line() +
  labs(
    title = "Normal Q-Q Plot",
    x = "Theoretical Quantile",
    y = "Standardized Residuals"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

shapiro.test(happiness$Residuals)
```

QQ plot is fine, but histogram doesn't look great. Shapiro says it's still 'ok' though. Maybe transform the data to get better normality.

```{r, Equal Variance}
# Errors have equal variance
autoplot(fit.happiness, which = 3, nrow = 1, ncol = 1) +
  labs(
    title = "Scale-Location",
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

Looks good.

```{r, Check for Influential Points}
# Check for influential points
autoplot(fit.happiness, which = 4, ncol = 1, nrow = 1) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

Nothing above (or even close) to 0.5, so we're good.

```{r, Multicollinearity, message=FALSE}
# Assess if multicollinearity is present
vif(fit.happiness)
mean(vif(fit.happiness))
```

GDP looks strongly correlated with a couple in the original scatter matrix, but no VIFs are above 10 and the mean is below 5, so we're good I guess?


# Interpret Coefficients of Chosen Model:

GDP: We expect an average increase in citizen happiness score for a country of ~0.8105 for every 1 unit (currency of the respective country?) increase in GDP.

Social Support: We expect an average increase in citizen happiness score for a country of ~1.0166 for every 1 point increase in social support score.

Healthy: We expect an average increase in citizen happiness score for a country of ~1.1414 for every 1 point increase in healthy life expectancy score.

Freedom: We expect an average increase in citizen happiness score for a country of 1.8458 for every 1 point increase in freedom to make life choices score.